---
title: "Translating NONMEM models to ferx"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Translating NONMEM models to ferx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

`ferxtranslate` converts NONMEM control stream files (`.ctl`) to ferx `.ferx`
format in one call. This vignette walks through the bundled warfarin example
from start to finish.

The translation chain uses `nonmem2rx` as a parser, so install it first:

```{r install}
install.packages(c("nonmem2rx", "ferxtranslate"))
```

## Step 1 -- Translate

Point `nm_to_ferx()` at a control stream file. If you also pass `output`, the
`.ferx` file is written immediately:

```{r translate}
library(ferxtranslate)

ctl    <- system.file("testmodels/nonmem/1cpt_oral.ctl", package = "ferxtranslate")
result <- nm_to_ferx(ctl, output = "warfarin.ferx")
```

`result` is a `ferx_translate_result` with three fields:

| Field | Contents |
|---|---|
| `$ferx_text` | The complete `.ferx` file as a character string |
| `$warnings` | Translation messages (INFO / WARN / ERROR prefixed) |
| `$unsupported` | Features present in the source that were not translated |

## Step 2 -- Read the output

```{r show}
cat(result$ferx_text)
```

The output for the warfarin 1-compartment oral model looks like:

```
# Translated from nonmem: 1cpt_oral.ctl

[parameters]
  theta TVCL(0.134, 0.001, 10.0)
  theta TVV(8.1, 0.1, 500.0)
  theta TVKA(1.0, 0.01, 50.0)

  omega ETA_CL ~ 0.07
  omega ETA_V ~ 0.02
  omega ETA_KA ~ 0.4

  sigma PROP_ERR ~ 0.01 (sd)

[individual_parameters]
  CL = TVCL * exp(ETA_CL)
  V = TVV * exp(ETA_V)
  KA = TVKA * exp(ETA_KA)

[structural_model]
  pk one_cpt_oral(cl=CL, v=V, ka=KA)

[error_model]
  DV ~ proportional(PROP_ERR)

[fit_options]
  method = focei
  maxiter = 500
  covariance = true
```

Always review:

- **Theta bounds** -- ferx always emits `(init, lower, upper)`. If the source
  lacked bounds, the translator uses `(init, 0, Inf)` and appends a WARN.
- **PK macro** -- inferred from the parameters defined in `$PK`. Check that
  `one_cpt_oral` (or `two_cpt_oral` etc.) matches your model intent.
- **Error model type** -- complex `$ERROR` blocks are classified heuristically;
  verify against the original.

## Step 3 -- Check warnings

```{r warnings}
result$warnings
result$unsupported
```

Common messages and what to do:

| Message | Action |
|---|---|
| `INFO | THETA X was FIXED` | ferx treats it as free; add a comment or pin post-fit |
| `WARN | Theta bounds not found` | Open the `.ferx` file and add realistic bounds |
| `WARN | complex $ERROR -- classified as proportional` | Verify error model against original |
| `ERROR | MIXTURE model detected` | Manual rewrite required; ferx does not support mixtures yet |

If `result$unsupported` is non-empty, the `.ferx` file has `# WARNING:` comments
at the affected locations so you can find them quickly.

## Step 4 -- Fit with ferx

Once you are happy with the `.ferx` file, pass it directly to `ferx_fit()`:

```{r fit}
library(ferx)
fit <- ferx_fit("warfarin.ferx", data = "warfarin.csv")
print(fit)
```

## Covariates

For models with covariates (weight, creatinine clearance), the allometric power
expressions pass through unchanged:

```{r cov}
ctl2   <- system.file("testmodels/nonmem/2cpt_oral_cov.ctl", package = "ferxtranslate")
result2 <- nm_to_ferx(ctl2)
cat(result2$ferx_text)
```

The `[individual_parameters]` section will contain lines such as:

```
CL = TVCL * (WT/70)^THETA_WT * (CRCL/100)^THETA_CRCL * exp(ETA_CL)
```

Check that the covariate column names in the `.ferx` file match the column
names in your dataset.

## ODE models

For `$DES` block models, ferxtranslate produces an `[odes]` section:

```{r ode}
ctl_ode <- system.file("testmodels/nonmem/ode_warfarin.ctl", package = "ferxtranslate")
result_ode <- nm_to_ferx(ctl_ode)
cat(result_ode$ferx_text)
```

The compartment names (`depot`, `central`) come from `$MODEL COMP=()` or from
the `d/dt()` call names. Verify that the `obs_cmt` in `[structural_model]`
points to the compartment that maps to DV in your dataset.

## nlmixr2 models

nlmixr2 model functions translate the same way via `nlmixr2_to_ferx()`:

```{r nlmixr2}
f_1cpt <- function() {
  ini({
    tvcl <- c(0.001, 0.134, 10.0)
    eta.cl ~ 0.07
    err.prop ~ 0.01
  })
  model({
    cl <- tvcl * exp(eta.cl)
    linCmt() ~ prop(err.prop)
  })
}
result_r <- nlmixr2_to_ferx(f_1cpt)
cat(result_r$ferx_text)
```

## What cannot be translated

See the `$unsupported` field and `# WARNING:` comments in the output for a
per-model list. The package-level catalogue:

```{r unsupported_table, echo = FALSE, eval = TRUE}
tbl <- data.frame(
  Feature = c(
    "3-compartment analytical",
    "Multiple DVIDs",
    "MIXTURE models",
    "R1 / D1 infusion rate/duration",
    "ALAG2 / ALAG3",
    "F2 / F3",
    "SIGMA BLOCK (correlated residuals)",
    "IOV with correlated kappas",
    "Categorical / count / TTE",
    "ADVAN5 (general linear, >3 cpts)"
  ),
  Status = rep("ERROR -- manual edit required", 10L),
  stringsAsFactors = FALSE
)
knitr::kable(tbl)
```

Each `ERROR` item produces a `# WARNING:` comment in the `.ferx` file at the
point where the feature would have appeared, so you know exactly what to fix.
